---
title: "Diversas formas do uso do tufte (panfleto)"
runningheader: "Tufte com R Markdown" # only for pdf output
subtitle: "" # only for html output
author: "Jeanne Franco"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: referencias.bib
link-citations: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

# Introdução

> *Nesse documento do estilo Tufte, mostrarei as diversas formas de uso
> que esse estilo permite. Para um texto introdutório, basta escrever
> diretamente com as citações que irão ser dispostas como notas e para
> os cabeçalhos, usamos a sintaxe Markdown de cerquilhas (\#). Veja o
> exemplo abaixo e observe o uso das funções do Tufte no arquivo do
> RMarkdown (Rmd).*

Os Direitos Humanos atribui que cada um de nós é um ser moral e racional
e que devemos viver com dignidade. Todos e todas deveríamos ter acesso
aos mesmos direitos, por isso eles são universais
[@InstitutoAurora2020].

De acordo com o Insituto Aurora, os direitos humanos incluem paz, saúde,
liberdade religiosa, educação, trabalho, autodeterminação, entre outros.

A Organização das Nações Unidas declara que todos os seres humanos
nascem livres e iguais em dignidade e direitos, dotados de consciência,
devem agir em sociedade com espírito de fraternidade. Garantir os
direitos humanos evita guerras, promove a paz e promove a
democracia.[^1]

[^1]: Ver Organização das Nações Unidas [Democracia e Direitos
    Humanos](https://brasil.un.org/pt-br/55783-democracia-e-direitos-humanos)

# Resultados

## Figuras nas margens

```{r pacotes, message=FALSE, warning=FALSE, include=FALSE}
library(rmarkdown)
library(tidyverse)
library(flextable)
library(gridExtra)
```

```{r dados, message=FALSE, warning=FALSE, include=FALSE}

dir_hum <- read.csv("human-rights-vdem.csv")
names(dir_hum)

dir_hum <- dir_hum %>%
  select(-Code, -civ_libs_vdem_low_owid, -civ_libs_vdem_high_owid) %>%
  filter(Entity == "Brazil",
  between(Year, 1930, 2021)) %>%
  rename(Index = civ_libs_vdem_owid) %>%
  view()
```

```{r tabela1, eval=FALSE, message=FALSE, warning=FALSE, comment='', include=FALSE}

### Carregar pacotes

library(rmarkdown)
library(tidyverse)
library(flextable)
library(gridExtra)

### Manipular dados 

dir_hum <- read.csv("human-rights-vdem.csv")

dir_hum <- dir_hum %>%
  select(-Code, -civ_libs_vdem_low_owid, -civ_libs_vdem_high_owid) %>%
  filter(Entity == "Brazil",
  between(Year, 1930, 2021)) %>%
  rename(Index = civ_libs_vdem_owid) 

### Tabela final de dados

dt <- dir_hum[1:15, 1:3]
dt %>%
  flextable(cwidth = 1.2) %>%
  set_caption(caption = "Dados parciais sobre direitos humanos no Brasil")
```

```{r tabela2, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

dir_hum1 <- dir_hum %>%
  group_by(Entity) %>%
  summarise(Mean = round(mean(Index), 2),
            sd = round(sd(Index), 2), n = n(),
            se = round(sd/sqrt(n), 2), 
            Median = round(median(Index), 2),
            Max = round(max(Index), 2), 
            Min = round(min(Index), 2),
            Range = Max - Min,
            Variance = round(var(Index), 2)) 

### Tabela da estatística descritiva

dir_hum1 %>%
  flextable(cwidth = 0.6) %>%
  set_caption(caption = "Estatística descritiva sobre direitos humanos no Brasil")
```

```{r gráficos, echo=FALSE, message=FALSE, warning=FALSE, fig.margin = TRUE, fig.cap = "Figura na margem do panfleto.", fig.width=3.5, fig.height=3.5, cache=TRUE}

### Histogramas

g1 <- ggplot(dir_hum, aes(x = Index, fill = Entity)) +
  geom_density(bins = 60, fill = "#d95f02") +
  labs(title = "",
       x = "Indice", y = "Densidade") +
  theme_light() +
  theme(legend.position = "none")

g2 <- ggplot(dir_hum, aes(x = Index, fill = Entity)) +
  geom_histogram(bins = 60, fill = "#d95f02") +
  labs(title = "", x = "Indice", y = "Frequencia") +
  theme_light() +
  theme(legend.position = "none")

### Box-plot

g3 <- ggplot(dir_hum, aes(y = Index, x = Entity)) +
  geom_boxplot(bins = 60, fill = "#d95f02") +
  stat_summary(fun.y = mean, geom = "point", shape = 20, 
               size = 7, color = "black") +
  labs(title = "", x = "Pais", y = "Indice") +
  theme_light() +
  theme(legend.position = "none")

### Linha

g4 <- ggplot(dir_hum, aes(x = Year, y = Index)) +
  geom_line(color = "#d95f02", size = 1.6) +
  labs(title = "", y = "Indice", x = "Tempo (anos)") +
  theme_light() +
  theme(legend.position = "none")

gridExtra::grid.arrange(g1, g2, g3, g4)
```

> *Note no arquivo RMarkdown o uso do `fig.cap` no chunk option para
> promover o título da figura. Você pode ajustar as proporções da figura
> usando `fig.width` e `fig.height` no chunk. Os ajustes são
> especificados em polegadas, e irão ser automaticamente escalados para
> a margem do panfleto.*

## Conteúdo para margem

> *Você pode incluir qualquer assunto na margem do panfleto nomeando um
> chunck `marginfigure`. Ao contrário dos blocos chunk de código do R
> ```` ```{r} ````, você escreve um chunk iniciando com
> ```` ```{marginfigure} ```` e então coloca o conteúdo do chunk.*

```{marginfigure}
library(rmarkdown)
library(tidyverse)
library(flextable)
library(gridExtra)
```

## Figuras de largura total

> *Você pode fazer com que a figura se organize por toda a página usando
> a opção de chunk: `fig.fullwidth = TRUE`.*

> *Outras opções relacionadas a esse tipo de figura podem ser usados com
> `fig.width`, `fig.cap`, `out.width` que adiciona um título e modifica
> o tamanho da altura e largura do gráfico.*

```{r fig-fullwidth, echo=FALSE, fig.cap="A full width figure.", fig.fullwidth=TRUE, fig.height=6, fig.width=8, message=FALSE, warning=FALSE, cache=TRUE}

### Histogramas

g1 <- ggplot(dir_hum, aes(x = Index, fill = Entity)) +
  geom_density(bins = 60, fill = "#d95f02") +
  labs(title = "",
       x = "Indice", y = "Densidade") +
  theme_light() +
  theme(legend.position = "none")

g2 <- ggplot(dir_hum, aes(x = Index, fill = Entity)) +
  geom_histogram(bins = 60, fill = "#d95f02") +
  labs(title = "", x = "Indice", y = "Frequencia") +
  theme_light() +
  theme(legend.position = "none")

### Box-plot

g3 <- ggplot(dir_hum, aes(y = Index, x = Entity)) +
  geom_boxplot(bins = 60, fill = "#d95f02") +
  stat_summary(fun.y = mean, geom = "point", shape = 20, 
               size = 7, color = "black") +
  labs(title = "", x = "Pais", y = "Indice") +
  theme_light() +
  theme(legend.position = "none")

### Linha

g4 <- ggplot(dir_hum, aes(x = Year, y = Index)) +
  geom_line(color = "#d95f02", size = 1.6) +
  labs(title = "", y = "Indice", x = "Tempo (anos)") +
  theme_light() +
  theme(legend.position = "none")

gridExtra::grid.arrange(g1, g2, g3, g4)
```

## Conteúdo na largura total do panfleto

> *Qualquer conteúdo pode ocupar toda a largura da página. O conteúdo
> deve estar cercado com `Div` e com a classe `fullwidth`.*

``` md
::: {.fullwidth}
Qualquer conteúdo aqui.
:::
```

*Aqui está um exemplo:*

::: fullwidth
**As três primeiras gerações dos direitos humanos se pautam nos
princípios da Revolução Francesa, a partir dessas três conhecidas
palavras: liberdade, igualdade e fraternidade.**
:::

## Figuras na coluna principal

> *Além das figuras nas margens e ocupando toda a página, você também
> pode incluir figuras restritas na coluna do panfleto.*

```{r fig-main, echo=FALSE, message=FALSE, warning=FALSE, fig.cap = "Figura na coluna principal.", cache=TRUE}

### Histogramas

g1 <- ggplot(dir_hum, aes(x = Index, fill = Entity)) +
  geom_density(bins = 60, fill = "#d95f02") +
  labs(title = "",
       x = "Indice", y = "Densidade") +
  theme_light() +
  theme(legend.position = "none")

g2 <- ggplot(dir_hum, aes(x = Index, fill = Entity)) +
  geom_histogram(bins = 60, fill = "#d95f02") +
  labs(title = "", x = "Indice", y = "Frequencia") +
  theme_light() +
  theme(legend.position = "none")

### Box-plot

g3 <- ggplot(dir_hum, aes(y = Index, x = Entity)) +
  geom_boxplot(bins = 60, fill = "#d95f02") +
  stat_summary(fun.y = mean, geom = "point", shape = 20, 
               size = 7, color = "black") +
  labs(title = "", x = "Pais", y = "Indice") +
  theme_light() +
  theme(legend.position = "none")

### Linha

g4 <- ggplot(dir_hum, aes(x = Year, y = Index)) +
  geom_line(color = "#d95f02", size = 1.6) +
  labs(title = "", y = "Indice", x = "Tempo (anos)") +
  theme_light() +
  theme(legend.position = "none")

gridExtra::grid.arrange(g1, g2, g3, g4)
```

# Notas laterais

> *Uma das mais proeminentes características do estilo panfleto é o
> extersivo uso de notas laterais. Existe nesse documento uma ampla
> margem para fazer uso das notas e pequenas figuras.*
>
> *Se você quiser colocar informações auxiliares na margem sem a marca
> da nota lateral (o número sobrescrito), você pode usar a função*
> `margin_note()` *do tufte em expressão inline do R.*
> `r margin_note("O objetivo desta pesquisa foi avaliar a evolução do índice de direitos humanos no Brasil entre os anos de 1930 e 2021.")`

# References

References can be displayed as margin notes for HTML output. For
example, we can cite R here [@R-base]. To enable this feature, you must
set `link-citations: yes` in the YAML metadata, and the version of
`pandoc-citeproc` should be at least 0.7.2. You can always install your
own version of Pandoc from <https://pandoc.org/installing.html> if the
version is not sufficient. To check the version of `pandoc-citeproc` in
your system, you may run this in R:

```{r eval=FALSE}
system2('pandoc-citeproc', '--version')
```

If your version of `pandoc-citeproc` is too low, or you did not set
`link-citations: yes` in YAML, references in the HTML output will be
placed at the end of the output document.

# Tables

You can use the `kable()` function from the **knitr** package to format
tables that integrate well with the rest of the Tufte handout style. The
table captions are placed in the margin like figures in the HTML output.

```{r}
knitr::kable(
  mtcars[1:6, 1:6], caption = 'A subset of mtcars.'
)
```

# Block Quotes

We know from the Markdown syntax that paragraphs that start with `>` are
converted to block quotes. If you want to add a right-aligned footer for
the quote, you may use the function `quote_footer()` from **tufte** in
an inline R expression. Here is an example:

> "If it weren't for my lawyer, I'd still be in prison. It went a lot
> faster with two people digging."
>
> `r tufte::quote_footer('--- Joe Martin')`

Without using `quote_footer()`, it looks like this (the second line is
just a normal paragraph):

> "Great people talk about ideas, average people talk about things, and
> small people talk about wine."
>
> --- Fran Lebowitz

# Responsiveness

The HTML page is responsive in the sense that when the page width is
smaller than 760px, sidenotes and margin notes will be hidden by
default. For sidenotes, you can click their numbers (the superscripts)
to toggle their visibility. For margin notes, you may click the circled
plus signs to toggle visibility.

# More Examples

The rest of this document consists of a few test cases to make sure
everything still works well in slightly more complicated scenarios.
First we generate two plots in one figure environment with the chunk
option `fig.show = 'hold'`:

```{r fig-two-together, fig.cap="Two plots in one figure environment.", fig.show='hold', cache=TRUE, message=FALSE}
p <- ggplot(mtcars2, aes(hp, mpg, color = am)) +
  geom_point()
p
p + geom_smooth()
```

Then two plots in separate figure environments (the code is identical to
the previous code chunk, but the chunk option is the default
`fig.show = 'asis'` now):

```{r fig-two-separate, ref.label='fig-two-together', fig.cap=sprintf("Two plots in separate figure environments (the %s plot).", c("first", "second")), cache=TRUE, message=FALSE}
```

You may have noticed that the two figures have different captions, and
that is because we used a character vector of length 2 for the chunk
option `fig.cap` (something like
`fig.cap = c('first plot', 'second plot')`).

Next we show multiple plots in margin figures. Similarly, two plots in
the same figure environment in the margin:

```{r fig-margin-together, fig.margin=TRUE, fig.show='hold', fig.cap="Two plots in one figure environment in the margin.", fig.width=3.5, fig.height=2.5, cache=TRUE}
p
p + geom_smooth(method = 'lm')
```

Then two plots from the same code chunk placed in different figure
environments:

```{r fig-margin-separate, fig.margin=TRUE, fig.cap=sprintf("Two plots in separate figure environments in the margin (the %s plot).", c("first", "second")), fig.width=3.5, fig.height=2.5, cache=TRUE}
knitr::kable(head(iris, 15))
p
knitr::kable(head(iris, 12))
p + geom_smooth(method = 'lm')
knitr::kable(head(iris, 5))
```

We blended some tables in the above code chunk only as *placeholders* to
make sure there is enough vertical space among the margin figures,
otherwise they will be stacked tightly together. For a practical
document, you should not insert too many margin figures consecutively
and make the margin crowded.

You do not have to assign captions to figures. We show three figures
with no captions below in the margin, in the main column, and in full
width, respectively.

```{r fig-nocap-margin, fig.margin=TRUE, fig.width=3.5, fig.height=2, cache=TRUE}
# a boxplot of weight vs transmission; this figure
# will be placed in the margin
ggplot(mtcars2, aes(am, wt)) + geom_boxplot() +
  coord_flip()
```

```{r fig-nocap-main, cache=TRUE}
# a figure in the main column
p <- ggplot(mtcars, aes(wt, hp)) + geom_point()
p
```

```{r fig-nocap-fullwidth, fig.fullwidth=TRUE, fig.width=10, fig.height=3, cache=TRUE}
# a fullwidth figure
p + geom_smooth(method = 'lm') + facet_grid(~ gear)
```

# Some Notes on Tufte CSS

There are a few other things in Tufte CSS that we have not mentioned so
far. If you prefer `r sans_serif('sans-serif fonts')`, use the function
`sans_serif()` in **tufte**. For epigraphs, you may use a pair of
underscores to make the paragraph italic in a block quote, e.g.

> *I can win an argument on any topic, against any opponent. People know
> this, and steer clear of me at parties. Often, as a sign of their
> great respect, they don't even invite me.*
>
> `r quote_footer('--- Dave Barry')`

We hope you will enjoy the simplicity of R Markdown and this R package,
and we sincerely thank the authors of the Tufte-CSS and Tufte-LaTeX
projects for developing the beautiful CSS and LaTeX classes. Our
**tufte** package would not have been possible without their heavy
lifting.

You can turn on/off some features of the Tufte style in HTML output. The
default features enabled are:

``` yaml
output:
  tufte::tufte_html:
    tufte_features: ["fonts", "background", "italics"]
```

If you do not want the page background to be lightyellow, you can remove
`background` from `tufte_features`. You can also customize the style of
the HTML page via a CSS file. For example, if you do not want the
subtitle to be italic, you can define

``` css
h3.subtitle em {
  font-style: normal;
}
```

in, say, a CSS file `my_style.css` (under the same directory of your Rmd
document), and apply it to your HTML output via the `css` option, e.g.,

``` yaml
output:
  tufte::tufte_html:
    tufte_features: ["fonts", "background"]
    css: "my_style.css"
```

There is also a variant of the Tufte style in HTML/CSS named "[Envisoned
CSS](https://github.com/nogginfuel/envisioned-css)". This style can be
used by specifying the argument `tufte_variant = 'envisioned'` in
`tufte_html()`[^2], e.g.

[^2]: The actual Envisioned CSS was not used in the **tufte** package.
    We only changed the fonts, background color, and text color based on
    the default Tufte style.

``` yaml
output:
  tufte::tufte_html:
    tufte_variant: "envisioned"
```

To see the R Markdown source of this example document, you may follow
[this link to
Github](https://github.com/rstudio/tufte/raw/main/inst/rmarkdown/templates/tufte_html/skeleton/skeleton.Rmd),
use the wizard in RStudio IDE
(`File -> New File -> R Markdown -> From Template`), or open the Rmd
file in the package:

```{r eval=FALSE}
file.edit(
  tufte:::template_resources(
    'tufte_html', '..', 'skeleton', 'skeleton.Rmd'
  )
)
```

This document is also available in
[Chinese](https://rstudio.github.io/tufte/cn/), and its `envisioned`
style can be found [here](https://rstudio.github.io/tufte/envisioned/).

```{r bib, include=FALSE}
# create a bib file for the R packages used in this document
knitr::write_bib(c('base', 'rmarkdown'), file = 'skeleton.bib')
```
